---
title: "Lab07"
author: "Yushi Li (A15639705)"
date: "2/8/2022"
output: pdf_document
---
# 1. PCA of UK food data
## Data import
```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)

# Q1. How many rows and columns are in your new data frame named x? What R 
# functions could you use to answer this questions?
dim(x)
```

## Checking the data
```{r}
# Preview the first 6 rows:
head(x)
# Remove row name for first row:
rownames(x) <- x[,1]
x <- x[,-1]
head(x)
# Checking dimensions again:
dim(x)
# An alternative approach to address incorrect row names:
x <- read.csv(url, row.names = 1)
head(x)


# Q2. Which approach to solving the ‘row-names problem’ mentioned above do you 
# prefer and why? Is one approach more robust than another under certain 
# circumstances? 

# A. I prefer the second method. It requires fewer lines of code and is also
# more robust due to the consistence of row.name=1 over x <- x[,1]. Calling
# x <- x[,1] multiple times would result in the loss of more than one row names.
```

## Spotting major differences and trends
```{r}
# Barplots don't really help much in terms of interpreting this set of data.
barplot(as.matrix(x), beside = T, col=rainbow(nrow(x)))


# Q3. Changing what optional argument in the above barplot() function results 
# in the following plot?

# A. Setting beside=F, or simply leave this argument out as it is F by default.
barplot(as.matrix(x), beside = F, col = rainbow(nrow(x)))


# Am I missing Q.4? I can't find it on the worksheet :(


# Q5. Generating all pairwise plots may help somewhat. Can you make sense of 
# the following code and resulting figure? What does it mean if a given point 
# lies on the diagonal for a given plot?
pairs(x, col = rainbow(10), pch = 16)

# A. I can somewhat make sense of the code and the resulting figure. The code
# makes one plot for each possible pair of regions (i.e. England vs. Scotland).
# The resulting figure show the relationships between the consumption of each 
# food type between a pair of regions. If a point is on the diagonal, it means 
# the consumption of that particular food is consistent with a linear
# relationship between two regions.


# Q6. What is the main differences between N. Ireland and the other countries 
# of the UK in terms of this data-set?

# A. Based on this particular set of data, an average person in N. Ireland 
# consumes significantly more fresh potatoes and less fresh fruits and alcoholic
# drinks than people from other countries of the UK.
```

## PCA to the rescue
```{r}
# Use the prcomp() PCA function 
pca <- prcomp( t(x) )
summary(pca)


# Q7. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points.
# Plot PC1 vs PC2
plot(pca$x[,1], pca$x[,2], xlab = "PC1", ylab = "PC2", xlim = c(-270,500))
text(pca$x[,1], pca$x[,2], colnames(x))


# Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document.
country_col <- c("orange", "magenta", "blue", "green")
plot(pca$x[,1], pca$x[,2], xlab = "PC1", ylab = "PC2", xlim = c(-270,500))
text(pca$x[,1], pca$x[,2], colnames(x), col = country_col)


# To calculate how much variation in the original data each PC accounts for:
v <- round( pca$sdev^2/sum(pca$sdev^2) * 100 )
v
# Results agree with prcomp(), with PC1, PC2, and PC3 accounting for 67%, 29%, 
# and 4% of total variation, respectively, as shown below using summary():
## the second row here...
z <- summary(pca)
z$importance

# Barplots of proportion of variance each PC accounts for:
barplot(v, xlab="Principal Component", ylab="Percent Variation")
```

## Digging deeper (variable loadings)
```{r}
## Lets focus on PC1 as it accounts for > 90% of variance 
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,1], las=2 )


# Q9. Generate a similar ‘loadings plot’ for PC2. What two food groups feature 
# prominently and what does PC2 mainly tell us about?
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,2], las=2 )

# A. Fresh_potatoes (again) and Soft_drinks feature most prominently in PC2, 
# with Fresh_potatoes again pushing N. Ireland in the positive direction and 
# with Soft_drinks pushing the rest of the countries in the negative direction 
# on the y-axis.
```

## Biplots
```{r}
## The inbuilt biplot() can be useful for small datasets 
biplot(pca)
```

# 2. PCA of RNA-seq data
## Import data
```{r}
url2 <- "https://tinyurl.com/expression-CSV"
rna.data <- read.csv(url2, row.names=1)
head(rna.data)


# Q10. How many genes and samples are in this data set?
dim(rna.data)

# A. There are 100 genes from 10 samples.
```

## Make the unpolished plot of PC1 and PC2 after PCA.
```{r}
pca2 <- prcomp(t(rna.data), scale=TRUE)
plot(pca2$x[,1], pca2$x[,2], xlab="PC1", ylab="PC2")
```
 
## Assess how much variance each PC accounts for:
```{r}
summary(pca2)

#make a quick scree plot:
plot(pca2, main="'Squick' plot")
```

## Try manually make the same scree plot:
```{r}
# Variance captured per PC 
pca.var <- pca2$sdev^2

# Percent variance is often more informative to look at 
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)
pca.var.per

# Make the scree plot
barplot(pca.var.per, main="Scree Plot", 
        names.arg = paste0("PC", 1:10),
        xlab="Principal Component", ylab="Percent Variation")
```

## Make the PCA plot more presentable:
```{r}
# Define and use a vector of colors for wt and ko samples
colvec <- colnames(rna.data)
colvec[grep("wt", colvec)] <- "blue"
colvec[grep("ko", colvec)] <- "green"

plot(pca2$x[,1], pca2$x[,2], col=colvec, pch=16,
     xlab=paste0("PC1 (", pca.var.per[1], "%)"),
     ylab=paste0("PC2 (", pca.var.per[2], "%)"))

text(pca2$x[,1], pca2$x[,2], labels = colnames(rna.data), pos=c(rep(4,5), rep(2,5)))
```

### Use ggplot2
```{r}
# Load ggplot2
library(ggplot2)

# Define a data frame containing the PCA data
df <- as.data.frame(pca2$x)

# Our first basic plot
ggplot(df) + 
  aes(PC1, PC2) + 
  geom_point()

# Add a 'wt' and 'ko' "condition" column
df$samples <- colnames(rna.data) 
df$condition <- substr(colnames(rna.data),1,2)

p <- ggplot(df) + 
        aes(PC1, PC2, label=samples, col=condition) + 
        geom_label(show.legend = FALSE)
p

# Polish the plot
p + labs(title="PCA of RNASeq Data",
       subtitle = "PC1 clealy seperates wild-type from knock-out samples",
       x=paste0("PC1 (", pca.var.per[1], "%)"),
       y=paste0("PC2 (", pca.var.per[2], "%)"),
       caption="BIMM143 example data") +
     theme_bw()
```

## Optional: find the 10 genes that contribute the most to variance in PC1:
```{r}
loading_scores <- pca2$rotation[,1]

## Find the top 10 measurements (genes) that contribute
## most to PC1 in either direction (+ or -)
gene_scores <- abs(loading_scores) 
gene_score_ranked <- sort(gene_scores, decreasing=TRUE)

## show the names of the top 10 genes
top_10_genes <- names(gene_score_ranked[1:10])
top_10_genes 
```
#
#
```{r}
sessionInfo()
```
