---
title: "Lab14"
author: "Yushi Li (A15639705)"
date: "3/2/2022"
output:
  pdf_document:
    latex_engine: xelatex
---

# Getting started
## Data overview
```{r}
# import vaccination data
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)

# view first and last date
head(vax$as_of_date)
tail(vax$as_of_date)
```
### Q1. What column details the total number of people fully vaccinated? 
persons_fully_vaccinated

### Q2. What column details the Zip code tabulation area? 
zip_code_tabulation_area

### Q3. What is the earliest date in this dataset? 
2021-01-05

### Q4. What is the latest date in this dataset? 
2022-03-01

```{r}
# use skimm
skimr::skim(vax)

# find out how many values r na
sum(is.na(vax$persons_fully_vaccinated))
sum(is.na(vax$persons_fully_vaccinated)) / nrow(vax) * 100
```
### Q5. How many numeric columns are in this dataset? 
10

### Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column? 
18338

### Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)? 
17.04%

### Q8. [Optional]: Why might this data be missing?
The data might not be collected daily in the 14-month period. It appears to be updated on a weekly basis.

## Working with dates
```{r}
# load the package
library(lubridate)

# check today's date
today()

# Specify that we are using the year-month-day format
vax$as_of_date <- ymd(vax$as_of_date)

# now we can do math with dates!
today() - vax$as_of_date[1]

# number of days that the dataset spans:
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]

# days since last update:
today() - vax$as_of_date[nrow(vax)]

# number of unique days in dataset:
length(unique(vax$as_of_date))
```
### Q9. How many days have passed since the last update of the dataset? 
1 day

### Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)? 
61 unique dates

# Working with ZIP codes
```{r}
# load package
library(zipcodeR)

# test run
geocode_zip('92037')

# calculate distance between 2 areas (in miles)
zip_distance('92037','92109')

# get census data from areas
reverse_zipcode(c('92037', "92109"))

# we can pull the data for ALL ZIP codes:
zipdata <- reverse_zipcode( vax$zip_code_tabulation_area )
```

# Focus on the San Diego area
## San Diego County at large
```{r}
# subset to San Diego county only areas using base R
sd <- vax[vax$county=="San Diego", ]
nrow(sd)

# do the same but with dplyr
library(dplyr)
sd <- filter(vax, county == "San Diego")
nrow(sd)

#subset over multiple criteria using dplyr
sd.10 <- filter(vax, county == "San Diego" &
                age5_plus_population > 10000)
```

### Q11. How many distinct zip codes are listed for San Diego County? 
```{r}
length(unique(sd$zip_code_tabulation_area))
head(sd)
```
There are 107 unique zip codes listed for San Diego County.

### Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset? 
```{r}
sd[which.max(sd$age12_plus_population), "zip_code_tabulation_area"]
```
92154

### Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2022-02-22”? 
```{r}
# select all San Diego county entries on as_of_date == “2022-02-22”
sd.20220222 <- filter(sd, as_of_date == "2022-02-22")

# skim
skimr::skim(sd.20220222)

# find overall average
mean(sd.20220222$percent_of_population_fully_vaccinated, na.rm = TRUE) * 100
```
The overall average is 70.42%.

### Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2022-02-22”?
```{r}
# use ggplot to make the figure
library(ggplot2)
ggplot(sd.20220222, aes(x = percent_of_population_fully_vaccinated)) +
  geom_histogram(bins = 12) +
  labs(title = "Histogram of Vaccination Rates across San Diego County", subtitle = "As of 2022-02-22", x = "Percent of Population Fully Vaccinated in a ZIP Code Area", y = "Count (ZIP code areas)")
```

## Focus on UCSD/La Jolla
```{r}
# define selection on ucsd/la jolla area by zip code 92037 and verify population
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

### Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area:
```{r}
ggplot(ucsd) +
  aes(x = as_of_date,
      y = percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x = "Date", y="Percent Vaccinated", title = "Vaccination rate for La Jolla CA 92037")
```

## Comparing to similar sized areas
```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2022-02-22")

head(vax.36)
```

### Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-02-22”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?
```{r}
# find mean vaccination rate
ca.mean <- mean(vax.36$percent_of_population_fully_vaccinated, na.rm = TRUE)

# add this as a straight horizontal line to plot
ggplot(ucsd) +
  aes(x = as_of_date,
      y = percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x = "Date", y="Percent Vaccinated", title = "Vaccination rate for La Jolla CA 92037") +
  geom_hline(aes(yintercept=ca.mean))
```

### Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-02-22”?
```{r}
fivenum(vax.36$percent_of_population_fully_vaccinated)
mean(vax.36$percent_of_population_fully_vaccinated)
```
Min:      0.3881090
1st Qu.:  0.6539015
Median:   0.7332750
Mean:     0.733385
3rd Qu.:  0.8027110
Max:      1.0000000

### Q18. Using ggplot generate a histogram of this data.
```{r}
ggplot(vax.36, aes(x = percent_of_population_fully_vaccinated)) +
  geom_histogram() +
  labs(x = "Percent Vaccinated") +
  xlim(0, 1)
```

### Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?
```{r}
# for 92040
vax %>% filter(as_of_date == "2022-02-22") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)

# for 92109
vax %>% filter(as_of_date == "2022-02-22") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)
```
Both are below the average vlaue calculated earlier (0.55 < 0.73, 0.72 < 0.73).

### Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.
```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)


ggplot(vax.36.all) +
  aes(x = as_of_date,
      percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="blue") +
  ylim(0,1) +
  labs(x="Date", y="Percent Vaccinated",
       title="Vaccination rate across California",
       subtitle="Only areas with a population above 36k are shown.") +
  geom_hline(yintercept = ca.mean, linetype = "dashed")
```

### Q21. How do you feel about traveling for Spring Break and meeting for in-person class afterwards?
I'd generally feel comfortable traveling and meeting in-person in CA afterwards based on current trends in vaccination rates. Some areas might require additional caution as the (full) vaccination rate remained below 50% as of now and show no significant increase in slope.