---
title: "Lab 6 R Functions"
author: "Yushi Li (A15639705)"
date: "2/5/2022"
output: pdf_document
---


```{r}
# Example input vectors to start with
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)


# defining function grade()
grade_pt <- function(x) {
  # set NA grades to 0 within each vector.
  x[is.na(x)] <- 0
  print(x)
  print(x[-which.min(x)])
  # calculating average grade while excluding the lowest score.
  overall <- mean(x[-which.min(x)])
  overall
}
grade_pt(student1)
grade_pt(student2)
grade_pt(student3)
# looks like it works!
```

Q1. Write a function grade() to determine an overall grade from a vector of 
student homework assignment scores dropping the lowest single score. If a 
student misses a homework (i.e. has an NA value) this can be used as a score to 
be potentially dropped. Your final function should be adquately explained with 
code comments and be able to work on an example class gradebook such as this one
in CSV format: “https://tinyurl.com/gradeinput” [3pts]

```{r}
#' Calculate the average grade of vector of scores dropping the lowest one.
#' Missing values are treated as zeros.
#'
#' @param x A numeric vector of scores.
#'
#' @return The average of the scores.
#' @export
#'
#' @examples
#' student <- c(100, NA, 90, 97, 80)
#' grade(student)


grade <- function(x) {
  # set NA grades to 0 within each vector
  x[is.na(x)] <- 0
  # calculating average grade while excluding the lowest score
  overall <- mean(x[-which.min(x)])
  overall
}

# importing gradebook
url <- "https://tinyurl.com/gradeinput"
gradebook <- read.csv(url, row.names = 1)
View(gradebook)

# fun grade() on gradebook
apply(gradebook, 1, grade)
```


Q2. Using your grade() function and the supplied gradebook, Who is the top
scoring student overall in the gradebook? [3pts]

```{r}
# assign grading results to grades
grades <- apply(gradebook, 1, grade)
# find the maximum in grades
which.max(grades)
# looks like student-18 has the highest grade overall.

```


Q3. From your analysis of the gradebook, which homework was toughest on 
students (i.e. obtained the lowest scores overall? [2pts]


```{r}
# method 1: using mean(), treating NAs as outliers (remove them).
which.min(apply(gradebook, 2, mean, na.rm = TRUE))
# based on the means, hw3 was the most challenging.

# method 2: using median(), again, treating NAs as outliers.
which.min(apply(gradebook, 2, median, na.rm = TRUE))
# based on the medians, hw2 was the most challenging.

# boxplot of results
boxplot(gradebook)
```


Q4. Optional Extension: From your analysis of the gradebook, which homework was 
most predictive of overall score (i.e. highest correlation with average grade 
score)? [1pt]

```{r}
# assign all NA to 0 on gradebook, save as masekd_gradebook
masked_gradebook <- gradebook
masked_gradebook[is.na(masked_gradebook)] <- 0
# assign the graded scores as results
results <- apply(gradebook, 1, grade)
# apply cor() to the processed gradebook
apply(masked_gradebook, 2, cor, x = results)
which.max(apply(masked_gradebook, 2, cor, x = results))
# hw5 has the highest correlation with average grade.
```



Q5. Make sure you save your Rmarkdown document and can click the “Knit” button 
to generate a PDF foramt report without errors. Finally, submit your PDF to 
gradescope. [1pt]
